Extended Inversion: when it works, and why

### 2020-12-08 repro/papers/skip/slides.tex
current
near done

Theorem Uh-Oh: If $d = F[m_*]w_{\lambda}$ (noise-free transient data) and $|m-m_*| > 2\lambda/r$, then
\[
J_{\rm FWI}[m;w_{\lambda},d] = \|d\|^2, \,\, \nabla_m J_{\rm FWI}[m;w_{\lambda},d] = 0
\]
}
\only<4>{
$J_{\rm FWI}[m;w_{\lambda},d]$, $d=F[m_*]w_{\lambda}$, $w_{\lambda}$ = Ricker wavelet center frequency = $1/\lambda$, $m_*=0.4$ s/km, $h=1 km$
\begin{center}
\includegraphics[height=4cm]{Fig/FWI.png}
\end{center}
Red: $\lambda=0.05$ (20 Hz), Blue: $\lambda = 0.25$ (40 Hz) [H. Chen et al. SEG 20]
}
\only<5>{
Postmortem:

@@

Rationale: more likely to be able to 
\begin{itemize}
\item find low residual (extended) models (``hug your data''),
\item explore useful update directions
\end{itemize}
}
\only<2>{
Cast: 
\begin{itemize}
\item model parameters $m$ (eg. $m$=slowness)
\item Wave operator $L[m]$ (eg. constant density acoustic op
\[
\left. L[m]=m^2\frac{\partial^2}{\partial t^2} - \nabla^2\right)
\]
\item Known (!) source field $q(\bx,t)$ (eg. point source $w(t)\delta(\bx-\bx_s)$
\item dynamic fields $u(\bx,t)$ solve $L[m]u=q$ (eg. $u=$ pressure
\item Sampling operator $P$ extracts data trace(s) (pressure or ...) from $u$
\item Modeling operator $F[m]=PL[m]^{-1}$
\end{itemize}
}
\only<3>{
FWI: given $d$, $q$, 
\[
\min_m \|Pu-d\|^2 \mbox{ subj to } L[m]u=q
\]
\[
\Leftrightarrow \min_m \|F[m]q-d\|^2
\]

@@

WRI: given $q,d$, 
\[
\min_{m,g} \|F[m]g+e[m]|^2 + \alpha^2 \|g|^2
\]
}
\only<5>{
Simplification 1:

Variable Projection (Golub \& Pereyra 73, 03): eliminate $g$ by solving quadratic minimization
\[
J_{\rm WRI}[m;q,d] = \min_{g} \|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2
\]
minimizer = $g[m;q,d]$, then
\[
=\|F[m]g[m;q,d]+e[m]|^2 + \alpha^2 \|g[m;q,d]\|^2
\]
$m$ minimizes $J_{\rm WRI}$ $\Leftrightarrow$ $(m,g[m;q,d])$ solves WRI problem 
}
\only<6>{
Simplification 2:

@@

Rank-Nullity Theorem: $g = n + F[m]^Ts$, $s \in $ range of $F$ = data, $n$ = non-radiating, {\em and}
$n$ and $F[m]^Ts$ are orthogonal
\[
\|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2 = \|F[m](F[m]^Ts + n)+e[m]\|^2 + \alpha^2\|F[m]^Ts + n\|^2
\]
\[
= \|F[m]F[m]^Ts+e[m]\|^2 + \alpha^2\|F[m]^Ts\|^2 + \alpha^2\|n\|^2
\]
so
\[
J_{\rm WRI}[m;q,d] = \min_s \|F[m]F[m]^Ts-e[m]\|^2 + \alpha^2\|F[m]^Ts\|^2 
\]
}
\only<7>{
Simplification 3:

@@

\begin{frame}\frametitle{Matched Source Extension}
\only<1>{
Back to FWI setting:
\[
F[m]w(t) = \frac{1}{4 \pi h}w(t-mh)
\]
A different extended inversion: add wavelet to unknowns

@@

Idea: after signature decon, wavelet should be {\em compact} $\sim$ nonzero only near $t=0$ - inspires {\em Matched Source Objective}
\[
\min_{m,w} \|F[m]w-d\|^2 + \|Aw\|^2, \,\,Aw(t)=tw(t)
\]
(S. 94, Plessix et al. 99, similar to Adaptive Waveform Inversion (AWI) Warner \& Guatsch 14)
}
\end{frame}
\end{document}

>>>>

Theorem Uh-Oh: If $d = F[m_*]w_{\lambda}$ (noise-free transient data) and $|m-m_*| > 2\lambda/h$, then
\[
J_{\rm FWI}[m;w_{\lambda},d] = \|d\|^2, \,\, \nabla_m J_{\rm FWI}[m;w_{\lambda},d] = 0
\]
}
\only<4>{
$J_{\rm FWI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = $1/\lambda$ Hz, $h=1$ km
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=5cm]{Fig/FWI.png}
\end{center}
\vspace{-0.5cm}
Red: $\lambda=0.05$ s (20 Hz), Blue: $\lambda = 0.025$ s (40 Hz) [H. Chen et al. SEG 20]
}
\only<5>{
Postmortem:

@@

Rationale: more likely to be able to 
\begin{itemize}
\item find low residual (extended) models (``hug your data''),
\item explore useful update directions
\end{itemize}
}
\only<2>{
Cast: 
\begin{itemize}
\item model parameters $m$ (eg. $m$=slowness)
\item Wave operator $L[m]$ (eg. constant density acoustic op $L[m]=m^2\partial_t^2 - \nabla^2$)
\item Known (!) source field $q(\bx,t)$ (eg. point source $w(t)\delta(\bx-\bx_s)$
\item dynamic fields $u(\bx,t)$ solve $L[m]u=q$ (eg. $u=$ pressure)
\item Sampling operator $P$ extracts data trace(s) (pressure or ...) from $u$
\item Modeling operator $F[m]=PL[m]^{-1}$
\end{itemize}
}
\only<3>{
FWI: given $d$, $q$, 
\[
\min_m \|Pu-d\|^2 \mbox{ subj to } L[m]u=q
\]
\[
\Leftrightarrow \min_m \|F[m]q-d\|^2
\]

@@

WRI: given $q,d$, 
\[
\min_{m,g} \|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2
\]
}
\only<5>{
Simplification 1:

Variable Projection (Golub \& Pereyra 73, 03): eliminate $g$ by solving quadratic minimization
\[
J_{\rm WRI}[m;q,d] = \min_{g} \|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2
\]
minimizer = $g[m;q,d]$, then
\[
=\|F[m]g[m;q,d]+e[m]\|^2 + \alpha^2 \|g[m;q,d]\|^2
\]
$m$ minimizes $J_{\rm WRI}$ $\Leftrightarrow$ $(m,g[m;q,d])$ solves WRI problem 
}
\only<6>{
Simplification 2:

@@

Rank-Nullity Theorem: $g = n + F[m]^Ts$, $s \in $ range of $F$ = data, $n$ = non-radiating, {\em and}
$n$ and $F[m]^Ts$ are orthogonal
\[
\|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2 = \|F[m](F[m]^Ts + n)+e[m]\|^2 + \alpha^2\|F[m]^Ts + n\|^2
\]
\[
= \|F[m]F[m]^Ts+e[m]\|^2 + \alpha^2\|F[m]^Ts\|^2 + \alpha^2\|n\|^2
\]
so
\[
J_{\rm WRI}[m;q,d] = \min_s \|F[m]F[m]^Ts-e[m]\|^2 + \alpha^2\|F[m]^Ts\|^2 
\]
NB: $g$ is space-time field, $s$ is data - vast memory reduction, practical 3D time domain (Yingst \& Wang SEG 16, Rizzuti et al. SEG 19)
}
\only<7>{
Simplification 3:

@@

\begin{frame}\frametitle{Matched Source Inversion}
\only<1>{
Back to FWI setting:
\[
F[m]w(t) = \frac{1}{4 \pi h}w(t-mh)
\]
A different extended inversion: add wavelet to unknowns

@@

Idea: after signature decon, wavelet should be {\em compact} $\sim$ nonzero only near $t=0$ - inspires {\em Matched Source Objective}
\[
\min_{m,w} \|F[m]w-d\|^2 + \alpha^2\|Aw\|^2, \,\,Aw(t)=tw(t)
\]
(S. 94, Plessix et al. 99, similar to Adaptive Waveform Inversion (AWI) Warner \& Guatsch 14)
}
\only<2>{
Variable Projection: 
\[
J_{\rm MSI}[m;d] = \min_w \|F[m]w-d\|^2 +\alpha^2 \|Aw\|^2
\]
Minimizer $w[m;d]$ solves normal equation
\[
(F[m]^TF[m] + \alpha^2A^TA) w = F[m]^T d
\]
Explicit calculation:
\[
F[m]^TF[m] + \alpha^2A^TA = \left(\frac{1}{(4\pi h)^2} + \alpha^2 t^2\right)I, \,\,F[m]^Td(t) = \frac{1}{4\pi h}d(t+mh)
\]
}
\only<3>{
Plug it in, get explicit formulas for everything - in particular, if data is noise-free, $d=F[m_*]w_{\lambda}$, then
\[
\nabla_m J_{\rm MSI}[m;d] = -h \alpha^2 \int \,dt\, \frac{t-(m-m_*)h}{(1+(4\pi h \alpha (t-(m-m_*)h))^2)^2}w_{\lambda}(t)^2
\]

Theorem Ah-Ha! If $d=F[m_*]w_{\lambda}$,
\[
\nabla_m J_{\rm MSI}[m;d] \left\{\begin{array}{c}
< 0 \mbox{ if }m < m_*-\lambda/h\\
> 0 \mbox{ if }m > m_*+\lambda/h
\end{array}
\right.
\]
$\Rightarrow$ if $m$ is stationary point of $J_{\rm MSI}[\cdot;d]$, then $|m-m_*| \le \lambda$.
}
\only<4>{
$J_{\rm MSI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = $1/\lambda$ Hz, $h=1$ km, $\alpha=1 s^{-1}km^{-1}$
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=5cm]{Fig/VPM.png}
\end{center}
\vspace{-0.5cm}
Red: $\lambda=0.05$ s (20 Hz), Blue: $\lambda = 0.025$ s (40 Hz) [H. Chen et al. SEG 20]
}
\only<5>{
$J_{\rm MSI} vs. J_{\rm FWI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = 40 Hz ($\lambda = 0.025$ s), $h=1$ km, $\alpha=1 s^{-1}km^{-1}$
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=5cm]{Fig/FWI-VPM-40.png}
\end{center}
\vspace{-0.5cm}
Red: MSI, Blue: FWI [H. Chen et al. SEG 20]
}
\only<6>{
$J_{\rm MSI} vs. J_{\rm FWI}$: $d=F[m_*]w_{\lambda}+n$, $n$ = 100\% RMS filtered stationary random noise, otherwise same
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=5cm]{Fig/FWI-VPM-noisy.pdf}
\end{center}
\vspace{-0.5cm}
Red: MSI, Blue: FWI [H. Chen]
}
\end{frame}

\end{document}
